---
title: "Getting Started with CellSweeper"
author:
  - name: Michael Totty
    affiliation: "Johns Hopkins Bloomberg School of Public Health"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Getting Started with CellSweeper}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment  = "#>",
    message  = TRUE,
    warning  = FALSE,
    fig.width  = 6,
    fig.height = 4
)
```

# Introduction

**CellSweeper** provides cluster-aware quality control (QC) for single-cell
resolution spatial transcriptomics data, including platforms such as Xenium,
CosMx, MERFISH, Stereo-seq, Slide-seq, and Visium HD.

It extends the [SpotSweeper](https://bioconductor.org/packages/SpotSweeper)
framework (Totty, Hicks & Guo, *Nature Methods* 2025) to single-cell
resolution, where SpotSweeper's core assumption --- that spatial neighbors are
biologically similar --- breaks down because adjacent cells can be entirely
different types.

## The three-level QC framework

1. **Level 1 --- Global pre-filtering** (`globalFilter()`): Remove obvious
   debris (near-zero counts, impossible segmentations).
2. **Level 2 --- Cluster-level QC** (`clusterCellTypes()` +
   `flagArtifactClusters()`): Cluster cells in expression space, then identify
   entire artifact clusters via multivariate pseudobulk outlier detection and
   spatial dispersion analysis.
3. **Level 3 --- Within-cluster spatial QC** (`clusterLocalOutliers()`):
   SpotSweeper-style local outlier detection restricted to same-cluster cells,
   applied to both transcriptomic and morphological metrics.

# Quick start with a synthetic dataset

For illustration in the vignette we build a small synthetic
`SpatialExperiment` with three spatially-coherent cell-type blobs plus a
handful of near-zero-count outliers per cluster. Real users would substitute
a Xenium/Visium HD/MERFISH object loaded via SpaceTrooper or another reader.

```{r build-synthetic, message = FALSE}
library(CellSweeper)
library(SpatialExperiment)
library(SummarizedExperiment)
suppressPackageStartupMessages({
    library(scuttle)
})

set.seed(1)
n_per <- 80
n_clusters <- 3
n_genes <- 200
n_cells <- n_per * n_clusters

centers <- matrix(c(0, 0, 100, 0, 50, 100), ncol = 2, byrow = TRUE)
coords  <- do.call(rbind, lapply(seq_len(n_clusters), function(i) {
    matrix(rnorm(n_per * 2, mean = 0, sd = 10), ncol = 2) +
        matrix(rep(centers[i, ], each = n_per), ncol = 2)
}))
colnames(coords) <- c("x_centroid", "y_centroid")

base_means <- c(5, 3, 1)
counts <- do.call(cbind, lapply(seq_len(n_clusters), function(i) {
    matrix(rpois(n_genes * n_per, lambda = base_means[i]),
        nrow = n_genes, ncol = n_per)
}))
# cluster-specific markers
for (i in seq_len(n_clusters)) {
    g <- ((i - 1) * 10 + 1):(i * 10)
    c <- ((i - 1) * n_per   + 1):(i * n_per)
    counts[g, c] <- counts[g, c] + 10
}
# inject 3 near-zero outliers per cluster
for (i in seq_len(n_clusters)) {
    out_idx <- ((i - 1) * n_per + 1):((i - 1) * n_per + 3)
    counts[, out_idx] <- 0
    counts[1:2, out_idx] <- 1
}
# mark mito genes
mito_idx <- (n_genes - 9):n_genes
rownames(counts) <- paste0("Gene", seq_len(n_genes))
rownames(counts)[mito_idx] <- paste0("MT-Gene", mito_idx)
colnames(counts) <- paste0("cell_", seq_len(n_cells))

spe <- SpatialExperiment(
    assays        = list(counts = as(counts, "CsparseMatrix")),
    spatialCoords = coords,
    sample_id     = rep("synthetic_sample", n_cells)
)
spe <- addPerCellQCMetrics(spe,
    subsets = list(mito = grep("^MT-", rownames(spe))))

# Pre-existing labels (we skip clustering in the quick-start example so
# the vignette runs in seconds; the per-step walk-through below shows
# how `clusterCellTypes()` would compute them from scratch).
spe$true_cluster <- rep(paste0("cluster_", seq_len(n_clusters)),
    each = n_per)
spe
```

```{r quickstart}
spe <- runCellSweeper(spe,
    use_existing   = "true_cluster",
    metrics        = c("sum", "detected", "subsets_mito_percent"),
    morpho_metrics = NULL,
    k              = 20,
    n_permutations = 50,
    verbose        = FALSE)

table(spe$cellsweeper_qc)

# Remove flagged cells for downstream analysis
spe_clean <- spe[, !spe$cellsweeper_qc]
ncol(spe_clean)
```

# Step-by-Step Workflow

## Step 1: Load data and compute QC metrics

CellSweeper operates on `SpatialExperiment` objects. Standard QC metrics
should be computed first using `scuttle::addPerCellQCMetrics()`.

For morphological metrics (`Area_um`, `log2AspectRatio`, `log2CountArea`),
we recommend preprocessing with
[SpaceTrooper](https://bioconductor.org/packages/SpaceTrooper) before running
CellSweeper. CellSweeper's defaults are aligned with SpaceTrooper's
`spatialPerCellQC()` output column names.

```{r load-data, eval=FALSE}
library(CellSweeper)
library(SpatialExperiment)
library(scuttle)

# Example: load Xenium data via SpaceTrooper
# library(SpaceTrooper)
# spe <- readXeniumSPE("/path/to/xenium/")
# spe <- spatialPerCellQC(spe)  # adds Area_um, log2AspectRatio, etc.

# Compute transcriptomic QC metrics
is_mito <- grep("^MT-", rownames(spe))
spe <- addPerCellQCMetrics(spe, subsets = list(mito = is_mito))
```

## Step 2: Global pre-filtering (Level 1)

Remove obvious debris: cells with near-zero counts or impossible
segmentations. If `Area_um` is present (from SpaceTrooper), area-based
filtering is also applied.

```{r global-filter, eval=FALSE}
spe <- globalFilter(spe,
    min_counts = 5,
    min_genes = 5,
    min_area = 10,      # if Area_um available
    max_area = 5000     # if Area_um available
)
```

If `Area_um` exists but `log2CountArea` does not, `globalFilter()` will
automatically compute transcript density (`CountArea` and `log2CountArea`)
for downstream use.

## Step 3: Cluster cells (Level 2a)

Cluster cells in expression space to group biologically similar cells.
This is a coarse clustering --- the goal is broad cell-type groupings,
not fine subtypes.

```{r cluster, eval=FALSE}
spe <- clusterCellTypes(spe,
    resolution = 0.5,
    regress_covariates = FALSE,
    exclude_mito = TRUE,
    exclude_ribo = TRUE,
    k.neighbors = 50
)

table(spe$cell_cluster)
```

If you already have cluster labels (e.g., from a reference-based annotation),
you can skip clustering:

```{r cluster-existing, eval=FALSE}
spe <- clusterCellTypes(spe, use_existing = "my_annotation_column")
```

### Multi-sample data

When your `SpatialExperiment` contains multiple samples (i.e., a `sample_id`
column in `colData` with more than one unique value), `clusterCellTypes()`
automatically applies [Harmony](https://portals.broadinstitute.org/harmony/)
batch correction on the PCA embedding before clustering. This removes
batch effects so that cluster labels are consistent across samples ---
"cluster_1" represents the same biology in every sample.

Harmony requires the `harmony` package (`install.packages("harmony")`).
If it is not installed, clustering proceeds on the uncorrected PCA with
an informative message.

The corrected embedding is stored as `reducedDim(spe, "HARMONY")` for
inspection. Downstream QC steps (artifact detection, spatial outlier
detection) still operate independently per sample.

```{r cluster-multisample, eval=FALSE}
# Multi-sample: Harmony runs automatically
spe <- clusterCellTypes(spe,
    samples = "sample_id",   # column identifying samples (default)
    resolution = 0.5,
    regress_covariates = FALSE
)

# Inspect the batch-corrected embedding
dim(reducedDim(spe, "HARMONY"))
```

## Step 4: Flag artifact clusters (Level 2b+c)

Identify entire clusters that are likely technical artifacts. Uses two
complementary approaches:

- **Pseudobulk multivariate outlier detection**: clusters with extreme
  QC metric profiles
- **Spatial dispersion analysis**: artifact clusters tend to be spatially
  scattered, unlike real cell types

```{r artifact-clusters, eval=FALSE}
spe <- flagArtifactClusters(spe,
    metrics = c("sum", "detected"),
    require_both = TRUE  # must be BOTH a QC outlier AND spatially incoherent
)

# How many clusters flagged?
table(spe$cluster_artifact_flag)
```

## Step 5: Within-cluster spatial QC (Level 3)

The core CellSweeper step: for each non-artifact cluster, detect cells
that are outliers relative to their same-cluster spatial neighbors.

```{r local-outliers, eval=FALSE}
spe <- clusterLocalOutliers(spe,
    metrics = c("sum", "detected"),
    morpho_metrics = c("Area_um", "log2AspectRatio", "log2CountArea"),
    k = 36,
    z_threshold = 3,
    combine = "union"
)

table(spe$cellsweeper_outlier)
```

## Step 6: Examine results

All results are stored in `colData(spe)`:

```{r results, eval=FALSE}
# Per-metric z-scores and outlier flags
head(colData(spe)[, c("sum_zscore", "sum_outlier",
    "detected_zscore", "detected_outlier")])

# Combined flags
table(spe$cellsweeper_outlier)   # within-cluster outliers
table(spe$cluster_artifact_flag) # artifact cluster members

# Remove flagged cells
spe_clean <- spe[, !spe$cellsweeper_outlier & !spe$cluster_artifact_flag]
```

# Visualization

CellSweeper provides several plotting functions (require `ggplot2`).

## Point-based plots

The default rendering uses points at cell centroid coordinates. These work
on any `SpatialExperiment`, regardless of whether polygon data is available.

```{r viz-points, fig.width = 5, fig.height = 4}
library(ggplot2)

# Spatial plot colored by library size, outliers circled in red
plotSpatialOutliers(spe, metric = "sum")

# Per-cluster QC distributions (violin + boxplot)
plotClusterQC(spe, metric = "sum")

# Spatial clusters with artifact clusters greyed out
plotSpatialDispersion(spe)
```

## Polygon-based visualization

When cell boundary polygons are available --- for example, from SpaceTrooper's
`addPolygonsToSPE()` which stores `sf` geometry in
`colData(spe)$polygons` --- CellSweeper's spatial plots can render filled cell
boundaries instead of points. This provides a much richer view of the tissue
and makes segmentation artifacts visually apparent.

**Requirements:** The `sf` package must be installed
(`install.packages("sf")`).

### Auto-detection

By default (`use_polygons = "auto"`), polygon rendering activates
automatically when the polygon column exists and `sf` is installed.
No code changes are needed --- the same plotting calls work with both
point and polygon data:

```{r polygon-auto, eval=FALSE}
# These calls auto-detect polygons and render cell boundaries
# If no polygons are present, they fall back to points seamlessly
plotSpatialOutliers(spe, metric = "sum")
plotSpatialDispersion(spe)
```

### Explicit control

You can force a specific rendering mode:

```{r polygon-explicit, eval=FALSE}
# Force polygon rendering (errors if polygons are unavailable)
plotSpatialOutliers(spe, metric = "sum", use_polygons = TRUE)

# Force point rendering even when polygons exist
plotSpatialOutliers(spe, metric = "sum", use_polygons = FALSE)
```

### Outlier visualization with polygons

When using polygons, `plotSpatialOutliers()` fills each cell polygon by the
QC metric value and overlays a colored border on outlier cells. This makes
it easy to spot which cells were flagged and why:

```{r polygon-outliers, eval=FALSE}
# Library size: outliers get red borders over the blue gradient fill
plotSpatialOutliers(spe, metric = "sum",
    colors = c("grey90", "navy"),
    outlier_color = "red",
    outlier_stroke = 0.8)

# Cell area: spot segmentation errors
plotSpatialOutliers(spe, metric = "Area_um",
    outlier_col = "Area_um_outlier",
    colors = c("white", "darkgreen"),
    outlier_color = "magenta")

# Transcript density: low-quality cells
plotSpatialOutliers(spe, metric = "log2CountArea",
    outlier_col = "log2CountArea_outlier",
    colors = c("grey95", "darkorange"))
```

### Cluster visualization with polygons

`plotSpatialDispersion()` fills polygons by cluster identity, with artifact
clusters rendered in grey. Thin borders between cells help delineate
individual cell boundaries:

```{r polygon-clusters, eval=FALSE}
# Default: thin grey borders to delineate cells within clusters
plotSpatialDispersion(spe)

# No borders (faster for large datasets)
plotSpatialDispersion(spe,
    polygon_border_color = NA)

# Thicker, darker borders for presentation figures
plotSpatialDispersion(spe,
    polygon_border_color = "grey20",
    polygon_border_width = 0.15)
```

### Custom polygon column

If your polygon data is stored under a different column name, specify it
with `polygon_col`:

```{r polygon-custom-col, eval=FALSE}
# SpaceTrooper may use "polygons.global" for global coordinates
plotSpatialOutliers(spe, metric = "sum",
    polygon_col = "polygons.global")
```

### Recommended SpaceTrooper + CellSweeper workflow

A complete workflow combining SpaceTrooper preprocessing with CellSweeper QC
and polygon visualization:

```{r full-workflow, eval=FALSE}
library(SpaceTrooper)
library(CellSweeper)
library(scuttle)
library(ggplot2)

# 1. Load platform-specific data and compute morphology metrics
spe <- readXeniumSPE("/path/to/xenium/output/")
spe <- spatialPerCellQC(spe)       # adds Area_um, log2AspectRatio, etc.
spe <- addPolygonsToSPE(spe)       # adds sf polygons to colData

# 2. Compute transcriptomic QC metrics
spe <- addPerCellQCMetrics(spe,
    subsets = list(mito = grep("^MT-", rownames(spe))))

# 3. Run CellSweeper pipeline
spe <- runCellSweeper(spe,
    morpho_metrics = c("Area_um", "log2AspectRatio", "log2CountArea"))

# 4. Visualize with polygon boundaries (auto-detected)
plotSpatialOutliers(spe, metric = "sum")
plotSpatialOutliers(spe, metric = "Area_um",
    outlier_col = "Area_um_outlier")
plotSpatialDispersion(spe)

# 5. Compare before/after with point fallback
plotSpatialOutliers(spe, metric = "sum", use_polygons = FALSE)  # points
plotSpatialOutliers(spe, metric = "sum", use_polygons = TRUE)   # polygons
```

# Session Info

```{r session-info}
sessionInfo()
```
