---
title: "Sketch-Based Clustering for Large Datasets"
author:
  - name: Michael Totty
    affiliation: "Johns Hopkins Bloomberg School of Public Health"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Sketch-Based Clustering for Large Datasets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    message = TRUE,
    fig.width = 7,
    fig.height = 5
)
```

# Introduction

Single-cell resolution spatial transcriptomics platforms routinely produce
datasets with hundreds of thousands to millions of cells. CellSweeper's
default clustering pipeline --- log-normalize, PCA, SNN graph, Leiden ---
scales well to tens of thousands of cells but can become a bottleneck at
larger scales, primarily due to the SNN graph construction and community
detection steps.

CellSweeper provides a **sketch-based clustering** approach to handle
these large datasets efficiently. The idea is simple:

1. Compute PCA on all cells (this scales linearly)
2. **Subsample** a representative subset of cells
3. Cluster only the subsample
4. **Project** cluster labels to all remaining cells via nearest-centroid
   assignment in PCA space

This can reduce clustering time by orders of magnitude while preserving
cluster quality, especially when combined with other performance options
like approximate PCA and approximate nearest neighbors.

In this vignette we demonstrate sketch clustering on a real dataset of
~77,000 nuclei from the
[spatialDLPFC](http://research.libd.org/spatialDLPFC/) project, available
through the `spatialLIBD` package.

> **Note on chunk evaluation.** Code chunks below are set
> `eval=FALSE` because the snRNA-seq dataset is ~1.1 GB and downloading,
> processing, and clustering it during package build is not feasible.
> The companion `getting_started.Rmd` vignette runs end-to-end on a
> compact synthetic dataset and exercises every exported function.

# Loading the data

We use the single-nucleus RNA-seq data from the spatialDLPFC project.
This dataset contains 77,604 nuclei from human dorsolateral prefrontal
cortex with expert-annotated cell types, making it an excellent test case
for sketch clustering:

- It has a realistic number of cells (~77k)
- Cell types are imbalanced (excitatory neurons dominate, but rare
  populations like microglia and endothelial/mural cells are present)
- Expert annotations allow us to evaluate whether sketch clustering
  preserves rare cell types

```{r load-data, eval=FALSE}
# library(CellSweeper)
devtools::load_all()
library(spatialLIBD)
library(SpatialExperiment)
library(SingleCellExperiment)
library(scuttle)
library(BiocNeighbors)
library(BiocSingular)
library(BiocParallel)

# Download the annotated snRNA-seq dataset
# (downloads ~1.1 GB on first use, cached for subsequent runs)
zip_path <- fetch_data("spatialDLPFC_snRNAseq")
tmpdir <- tempdir()
unzip(zip_path, exdir = tmpdir)
sce <- HDF5Array::loadHDF5SummarizedExperiment(
    file.path(tmpdir, "sce_DLPFC_annotated")
)

# 77,604 cells x 36,601 genes
dim(sce)
table(sce$cellType_broad_hc)
#  Astro EndoMural Excit Inhib Micro Oligo   OPC Ambiguous
#   3979      2157 24809 11067  1601 10894  1940    21157
```

The dataset has 8 broad cell types with highly imbalanced frequencies ---
excitatory neurons make up 32% of cells while microglia are only 2%.
There are also 21,157 "Ambiguous" cells that could not be confidently
assigned to a single type.

Since this is snRNA-seq data (no tissue coordinates), we convert to a
`SpatialExperiment` with placeholder spatial coordinates. Clustering
operates entirely in expression space and does not use spatial
coordinates.

```{r convert-spe, eval=FALSE}
# Convert SingleCellExperiment to SpatialExperiment
# Realize HDF5-backed counts into an in-memory sparse matrix
counts_mat <- as(counts(sce), "dgCMatrix")

coords <- matrix(rnorm(ncol(sce) * 2), ncol = 2,
    dimnames = list(colnames(sce), c("x", "y")))
spe <- SpatialExperiment(
    assays = list(counts = counts_mat),
    spatialCoords = coords,
    colData = colData(sce)
)

# Ensure QC metrics are present
spe <- addPerCellQCMetrics(spe,
    subsets = list(mito = grep("^MT-", rownames(spe))))

# Store expert annotations for later comparison
spe$expert_broad <- spe$cellType_broad_hc
spe$expert_fine <- spe$cellType_hc
```

# The problem with uniform random sampling

The naive approach --- `sample(n_cells, subsample_size)` --- draws cells
uniformly at random. This works adequately when cell types are roughly
balanced, but our DLPFC dataset is highly imbalanced. Uniform sampling of
20,000 cells from 77,604 would yield only ~330 microglia (2% of 20,000),
potentially too few for the clustering algorithm to resolve them as a
distinct group.

# Leverage score sampling

CellSweeper's default sketch approach uses **leverage score sampling**
instead of uniform random sampling. Leverage scores come from randomized
numerical linear algebra and measure how much each cell contributes to
the low-rank PCA structure of the dataset.

## How it works

After PCA, each cell is represented as a point in PC space. The
**statistical leverage** of cell $i$ is:

$$\ell_i = \sum_{j=1}^{d} u_{ij}^2$$

where $U$ is the matrix of left singular vectors (the PCA loadings
normalized to unit variance per component) and $d$ is the number of PCs.

Cells in dense, common regions of PC space have low leverage --- they are
well-represented by many nearby cells and contribute little unique
information. Cells in sparse regions (rare cell types, transitional states,
outliers) have high leverage --- they disproportionately define the PC
directions and would be poorly represented if missed.

Sampling with probability proportional to leverage scores ensures that:

- **Rare populations are preserved.** The ~1,600 microglia and ~2,200
  endothelial/mural cells occupy unique regions of PC space and will
  have elevated leverage scores, so they are sampled at a higher rate
  than their small frequencies would suggest under uniform sampling.
- **Common populations remain well-covered.** Even with leverage-biased
  sampling, the sheer number of excitatory neurons ensures adequate
  representation.
- **The sketch faithfully represents the full PCA structure.** This is
  the key theoretical guarantee --- leverage score sampling preserves the
  spectral properties of the data matrix.

## Implementation in CellSweeper

Under the hood, CellSweeper:

1. Runs PCA on all cells
2. Scales each PC column to unit variance (recovering the left singular
   vectors from the PC scores)
3. Computes leverage as the row sum of squares
4. Normalizes to probabilities and samples without replacement
5. Clusters the subsample via Leiden (or mini-batch k-means)
6. Computes cluster centroids in PCA space
7. Assigns every cell to its nearest centroid via `BiocNeighbors::queryKNN()`

# Comparison: leverage vs. random vs. full clustering

Now we compare three approaches on our 77k-cell dataset: full clustering
(all cells, no subsample), sketch clustering with leverage sampling
(20k subsample), and sketch with random sampling (20k subsample).

```{r comparison, eval=FALSE}
# Full clustering (no subsample) --- baseline
system.time(
    spe_full <- clusterCellTypes(spe,
        resolution = 0.5,
        regress_covariates = FALSE,
        exclude_mito = TRUE
    )
)

# Sketch clustering with leverage sampling (default)
system.time(
    spe_lev <- clusterCellTypes(spe,
        subsample = 20000,
        subsample_method = "leverage",
        resolution = 0.5,
        regress_covariates = FALSE,
        exclude_mito = TRUE
    )
)

# Sketch clustering with random sampling
system.time(
    spe_rand <- clusterCellTypes(spe,
        subsample = 20000,
        subsample_method = "random",
        resolution = 0.5,
        regress_covariates = FALSE,
        exclude_mito = TRUE
    )
)
```

## Evaluating cluster quality against expert annotations

We can compare the discovered clusters against the expert-annotated cell
types to assess how well each method recovers the known biology:

```{r evaluate, eval=FALSE}
# Cross-tabulate discovered clusters vs expert cell types
# Full clustering
table(spe_full$cell_cluster, spe_full$expert_broad)

# Leverage sketch
table(spe_lev$cell_cluster, spe_lev$expert_broad)

# Random sketch
table(spe_rand$cell_cluster, spe_rand$expert_broad)
```

The key question is whether rare populations (Micro, EndoMural, OPC) are
resolved as distinct clusters. Leverage sampling increases the chance that
these rare cells are included in the sketch, giving the clustering algorithm
enough representatives to identify them as separate groups.

On balanced datasets, both sampling methods produce similar results. The
advantage of leverage sampling becomes apparent when rare populations are
present --- leverage sampling resolves them as distinct clusters more
reliably.

## Visualizing clusters with UMAP

We can visualize the clustering results on a shared UMAP embedding to
compare how well each method resolves cell populations. Since all three
methods share the same PCA (computed on all cells), we compute UMAP once
and overlay each set of cluster labels.

```{r umap-compute, eval=FALSE}
library(scater)
library(ggplot2)

# Compute UMAP from the shared PCA (stored by clusterCellTypes)
spe_full <- scater::runUMAP(spe_full, dimred = "PCA")
umap_df <- as.data.frame(reducedDim(spe_full, "UMAP"))
colnames(umap_df) <- c("UMAP1", "UMAP2")

# Attach cluster labels from each method
umap_df$full_cluster <- spe_full$cell_cluster
umap_df$leverage_cluster <- spe_lev$cell_cluster
umap_df$random_cluster <- spe_rand$cell_cluster
umap_df$expert <- spe_full$expert_broad
```

```{r umap-plot, eval=FALSE, fig.width=10, fig.height=8}
# Expert annotations (ground truth)
p_expert <- ggplot(umap_df, aes(UMAP1, UMAP2, color = expert)) +
    geom_point(size = 0.1, alpha = 0.3) +
    ggtitle("Expert annotations") +
    guides(color = guide_legend(override.aes = list(size = 3, alpha = 1))) +
    theme_minimal()

# Full clustering (all 77k cells)
p_full <- ggplot(umap_df, aes(UMAP1, UMAP2, color = full_cluster)) +
    geom_point(size = 0.1, alpha = 0.3) +
    ggtitle("Full clustering (all cells)") +
    theme_minimal() + theme(legend.position = "none")

# Sketch: leverage sampling (20k subsample)
p_lev <- ggplot(umap_df, aes(UMAP1, UMAP2, color = leverage_cluster)) +
    geom_point(size = 0.1, alpha = 0.3) +
    ggtitle("Sketch: leverage sampling (20k)") +
    theme_minimal() + theme(legend.position = "none")

# Sketch: random sampling (20k subsample)
p_rand <- ggplot(umap_df, aes(UMAP1, UMAP2, color = random_cluster)) +
    geom_point(size = 0.1, alpha = 0.3) +
    ggtitle("Sketch: random sampling (20k)") +
    theme_minimal() + theme(legend.position = "none")

# Arrange in a 2x2 grid
gridExtra::grid.arrange(p_expert, p_full, p_lev, p_rand, ncol = 2)
```

The UMAP plots reveal how each clustering approach partitions the
expression landscape. The expert annotations provide the ground truth,
while the full and sketch-based clusterings should show similar broad
structure. Differences between leverage and random sketch are most visible
in small, isolated UMAP islands corresponding to rare populations.

# The `leverage_alpha` mixing parameter

Pure leverage sampling can occasionally over-concentrate on extreme
outliers at the expense of the "bulk" of common cell types. The
`leverage_alpha` parameter controls a mixture between leverage-biased
and uniform sampling:

$$p_i = \alpha \cdot \frac{\ell_i}{\sum_j \ell_j} + (1 - \alpha) \cdot \frac{1}{n}$$

- `leverage_alpha = NULL` (default): pure leverage scores
- `leverage_alpha = 0.9`: 90% leverage + 10% uniform
- `leverage_alpha = 0.5`: equal mixture

```{r alpha-example, eval=FALSE}
# Mix leverage with uniform for more balanced coverage
spe_mixed <- clusterCellTypes(spe,
    subsample = 20000,
    leverage_alpha = 0.9,    # 90% leverage + 10% uniform
    resolution = 0.5,
    regress_covariates = FALSE,
    exclude_mito = TRUE
)
table(spe_mixed$cell_cluster, spe_mixed$expert_broad)
```

For most datasets, pure leverage (`leverage_alpha = NULL`) works well.
Consider adding a uniform component (`0.5`--`0.9`) if you find that
the sketch over-represents very small outlier populations at the expense
of resolving subtypes within the dominant population.

# Multi-sample data and Harmony

When working with multi-sample datasets, `clusterCellTypes()` automatically
applies [Harmony](https://portals.broadinstitute.org/harmony/) batch
correction on the PCA embedding before clustering. In the sketch workflow,
Harmony runs on the **full** PCA matrix (all cells), and then the subsample
is drawn from the corrected embedding. This ensures batch correction is
learned from the complete dataset, not just the sketch.

```{r harmony-sketch, eval=FALSE}
# Multi-sample + sketch: Harmony runs on full PCA, then subsample
spe <- clusterCellTypes(spe,
    subsample = 20000,
    samples = "sample_id",  # Harmony uses this for batch correction
    resolution = 0.5,
    regress_covariates = FALSE
)

# Both embeddings are available
dim(reducedDim(spe, "PCA"))      # uncorrected
dim(reducedDim(spe, "HARMONY"))  # batch-corrected (used for clustering)
```

# Combining with other performance options

The sketch approach can be combined with CellSweeper's other performance
features for maximum speed on very large datasets:

```{r full-fast, eval=FALSE}
library(BiocSingular)
library(BiocNeighbors)

spe_fast <- clusterCellTypes(spe,
    # Sketch clustering
    subsample = 20000,
    subsample_method = "leverage",

    # Approximate PCA (IRLBA instead of exact SVD)
    BSPARAM = IrlbaParam(),

    # Approximate nearest neighbors (HNSW for SNN graph)
    BNPARAM = HnswParam(),

    # Multi-threaded graph construction + Leiden
    num_threads = 4,

    resolution = 0.5,
    regress_covariates = FALSE,
    exclude_mito = TRUE
)
```

Or through the full pipeline wrapper:

```{r wrapper-fast, eval=FALSE}
library(BiocParallel)
library(BiocSingular)
library(BiocNeighbors)

spe <- runCellSweeper(spe,
    # Sketch clustering
    subsample = 20000,
    subsample_method = "leverage",

    # Fast PCA
    BSPARAM = IrlbaParam(),

    # Approximate NN (used in clustering, outlier detection,
    # and spatial homogeneity)
    BNPARAM = HnswParam(),

    # Parallel permutation testing + per-cluster outlier detection
    BPPARAM = MulticoreParam(4),

    # Multi-threaded Leiden
    num_threads = 4
)
```

## Performance option summary

| Parameter | Default | Fast option | Affects |
|:----------|:--------|:------------|:--------|
| `subsample` | `NULL` (all cells) | `20000` | Clustering |
| `subsample_method` | `"leverage"` | `"leverage"` | Clustering |
| `BSPARAM` | `NULL` (exact SVD) | `IrlbaParam()` | PCA |
| `BNPARAM` | `NULL` (exact kNN) | `HnswParam()` | SNN graph, spatial outliers, homogeneity |
| `BPPARAM` | `NULL` (serial) | `MulticoreParam(4)` | Regression, permutations, per-cluster QC |
| `num_threads` | `1` | `4` | SNN graph + Leiden |
| `cluster_method` | `"leiden"` | `"mbkmeans"` | Clustering algorithm |

# Using pre-existing labels

If you already have cluster labels from an external source --- such as a
reference-based annotation (SingleR, Azimuth), a published cell atlas, or
a previous analysis --- you can skip clustering entirely. For our DLPFC
dataset, we can use the expert annotations directly:

```{r use-existing, eval=FALSE}
# Use the expert broad cell type annotations directly
spe_annotated <- clusterCellTypes(spe,
    use_existing = "expert_broad"
)
table(spe_annotated$cell_cluster)

# Or in the full pipeline
spe <- runCellSweeper(spe, use_existing = "expert_broad")
```

This bypasses all clustering steps (normalization, PCA, graph
construction) and directly uses the provided labels for the downstream
artifact detection and within-cluster spatial QC.

# When to use sketch clustering

**Use sketch clustering when:**

- Your dataset has more than ~50,000--100,000 cells
- Clustering is a bottleneck in your pipeline
- You are running CellSweeper on multiple samples iteratively

**Skip sketch clustering when:**

- Your dataset has fewer than ~50,000 cells (standard clustering is fast
  enough)
- You need the most precise possible cluster boundaries (sketch
  introduces a small approximation)
- You already have cluster labels (`use_existing` is faster than any
  sketch)

As a rule of thumb, set `subsample` to 30,000--100,000. This provides
enough cells for robust cluster detection while keeping runtime
manageable. The exact number matters less than you might expect ---
leverage score sampling ensures that even a relatively small subsample
captures the diversity of the full dataset.

# Session Info

```{r session-info}
sessionInfo()
```
